---
title: "Lab 8"
format: html
---


#Breast Cancer Project
Today we are going to explore some data from the Univeristy of Wisconsin Cancer Center on Breast biosy data.

```{r}
wisc.data <-read.csv("/Users/JoshTran/Desktop/BIMM 143/Week 5/WisconsinCancer.csv", row.name = 1)
head(wisc.data)
```

> Q1. How many patients samples are in this dataset?


```{r}
nrow(wisc.data)
```


There are `r nrow(wisc.data)` patients in this data set.

> Q2. How many cancer (M) and non cancer (B) samples are there?

```{r}
table(wisc.data$diagnosis)
```


```{r}
head(wisc.data)
```

```{r}
diagnosis <- as.factor(wisc.data$diagnosis)
#diagnosis
```

```{r}
head(diagnosis)
```



Now exclude the diagnosis cloumn from the data

```{r}
wisc.data <- wisc.data[,-1]
```

> Q3. How many "dimensions","Variables", "columns", are in this dataset?

```{r}
ncol(wisc.data)
```


#Principal Component Analysis (PCA)

To perform PCA in R we can use `prcomp()` function.
it takes as input a numeric dataset and optional `scale = FALSE/TRUE` argument.

We generally always want to set the `scale=TRUE` but lets make sure by checking if the mean and standard deviation values are different across these 30 columns.

```{r}
round(colMeans(wisc.data) )
```


```{r}
pca <- prcomp(wisc.data, scale = TRUE)
summary(pca)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?
0.4427

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?
0.72636

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?
PC7




```{r}
attributes(pca$x)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

This is a hot mess of a plot and we will need to generate our own plots to make sense of this PCA result.



```{r}
plot(pca$x[,1], pca$x[,2], col= diagnosis)
```

```{r}
# Create a data.frame for ggplot
df <- as.data.frame(pca$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col= diagnosis) + 
  geom_point()

```

> Q8. How much variance is captured in the top 3 PCs.

They capture 72.636% of the total variance.


> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? 

This tells us how much this original feature contributes to the first PC.

```{r}
pca$rotation["concave.points_mean",1]
```




```{r}
attributes(pca)
```

#Combine PCA results with clustering

We can use our new PCA variables (i.e the scores along the PCs contained in R `pca$x` ) as input for other methods such as clustering

```{r}
#Hclust needs a distance matrix as input

d <- dist(pca$x [,1:3] )

hc <- hclust(d, method="ward.D2")
plot(hc)

```

> Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

h = 35.5

To get our cluster membership vector we can use the `cutree()` function and specify a height (`h`) or number of groups(`k`).

```{r}
grps <- cutree(hc, h=35.5)
table(grps)
```


I want to find out how many diagnosis "M" and "B" are in each grp?



```{r}
table(diagnosis)
```

> Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?


```{r}
table(diagnosis, grps)
```


> Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

We can also plot our results using our clustering vector `grps` 

```{r}
plot(pca$x[,1:2], col=grps)
```



```{r}
# Create a data.frame for ggplot
df <- as.data.frame(pca$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col= grps) + 
  geom_point()

```

> Q. Q15. What is the specificity ( TN/(TN+FN) ) and sensitivity( TP/(TP+FN) ) of our current results?

Specificity= (333/(333+33)) = .91 Sensitivity = (179/(179+33)) = .84

```{r}

```

## Predicition 

> Q16. Which of these new patients should we prioritize for follow up based on your results?

Patient 2 should follow up.





